Leadership across cultures and contexts: an exploratory analysis

### 2020-05-13 leadership_across_cultures_contexts.Rmd
Much work on discussion section. Added upset plot.

Hunter-gatherers and pastoralists were more likely to provide evidence for high status leaders, as were South America and East Eurasia. Female leaders were also more likely to be identified as high status. Somewhat counter-intuitively, the proportion of text records indicating high status leaders was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus).

@@

```{r shamanism, fig.width=10, fig.cap="Coefficients of logistic elasticnet regression models of evidence for *Shamanism*. A: Coefficients indicate the words whose frequencies in each text record best predicted evidence for *Shamanism* in each text record. B: Non-zero coefficients of all quality and function variables that best predicted evidence for *Shamanism*. Both models used the lasso penalty ($\\alpha=1$) and chose $\\lambda_{1se}$ by cross-validation (1 SE from $\\lambda_{min}$). Red values are positive predictors; purple values are negative predictors. Variables with coefficents = 0 not displayed. X-axes are odds ratios on a log 10 scale."}
plot_shamanism_text <- plot_shamanism_text + scale_x_log10()

@@

Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported, suggesting leadership to be a high-risk, high-return strategy (see Figure \@ref(fig:varsupportcb)). The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Followers were not reported to receive mating benefits, relative to leaders or other benefit types. Our results cannot determine if the benefits to leaders and followers outweighed their costs, but they do shed light on the social and biological pressures associated with leadership and social hierarchy. For both leaders and followers, individuals stood to lose what they were able to gain. See Figure \@ref(fig:varsupportcb).

@nm09-056 [p. 183] describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous female leader among the Iroquois, whose life exemplified the risks to both leaders and followers:

@@

All evolutionary models of leadership propose that the fitness benefits of leading outweigh the costs [e.g., @van_vugt_leadership_2008; @kantner_identifying_2010]. The relative abundance of evidence for leader and follower benefits vs. costs ($OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$ and $OR=`r signif(follower_bc_ratio$benefit_cost_OR, 2)`$, resp.), while not providing evidence of net fitness benefits, is nevertheless encouraging, especially since leaders and followers both had more evidence of mating and territory benefits than costs, which are key contributors to fitness, and leaders also had more evidence for inclusive fitness, material and social benefits, relative to costs. See Figure \@ref(fig:benefitcostratio).

@@

Shamanism appears to be a distinct form of leadership that combines a strategy of inducing fear, similar to the dominance strategy, but is based on knowledge and expertise, similar to the prestige strategy. The *Competencies* feature included supernatural and feared dimensions (Figure \@ref(fig:cluster)). Our *Shamanism* variable (which was not one of our original leadership dimensions) was predicted by words that evoke dominance and prestige, such as *supernatural*, *power*, *fear*, and *knowledge* (Figure \@ref(fig:shamanism)A), and by four other leadership dimensions that also overlap with the dominance and prestige constructs: *Feared*, *Medicinal functions*, *Ritual functions*, and *Experience/accomplished*. Shamans often provide valued public goods, have special expertise and abilities, are feared, and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate superhuman abilities [@singh_cultural_2017]. @ng06-054 describes shaman leaders among the Ojibwa:

@@

<!-- DISCUSSES PROSOCIALITY FEATURE -->
**Prosocial norms**. The prosociality feature of leader qualities, including generosity, interpersonal and oratory skills, and charisma as well as the role of cultural norms in shaping leadership styles, including proper behavior and having a favorable personality (Figure \@ref(fig:cluster)), generally corresponds to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @grabo_charisma_2017; @howell_role_2005]. The importance of leader prosociality also supports the @henrich_big_2015 Big-man mechanism by which the positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

<!-- DISCUSSES SUCCESS FEATURE -->
**Leader success**. Leaders, probably as a consequence of their capacities to provide prosocial benefits (e.g., resolving conflicts and being generous) and/or impose social costs (e.g., punishing and aggressiveness), are also characterized by their material, social, and reproductive success (bottom cluster Figure \@ref(fig:cluster)A), which is consistent with much of the theoretical literature focused on associated outcomes of leadership and social hierarchy [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013]. Several of these measures of success often co-occur. Among the Trobriand horticulturalists in Papua New Guinea, @ol06-037 [p. 47] describes relationships between polygynous marriages, increased wealth, and widespread social contacts:

>Chiefs managed unusually large quantities of resources and it is agreed that the basis of such wealth was polygamy. Rank was quite directly expressed in the number of wives. The chief of Omarakana had 16 wives in Malinowski's time, while his predecessor Enamakala had 19 in 1893 (Annual Report 1893-4:19). In 1950 despite long-established missionary activity, there were still 13 (Powell 1960: 136). In traditional Trobriand society it seems that all leaders of village clusters of guyau status became polygamists. An emerging leader might also expect to be given wives by traditional allies outside the cluster while he might demand them from traditional rivals of subordinate subclans (Powell 1960:136)....

<!-- DISCUSSES BOTH MANAGEMENT FEATURES -->
**Leaders are managers**. There is a history within managerial studies among WEIRD populations of contrasting *leaders* from *managers*. @zaleznik_managers_1977, for example, suggested "managers" seek systematicity, organization, and stability, whereas "leaders" are willing to accept a lack of control and creatively navigate uncertainty to produce novel insight and appeal to the emotions of followers. Political anthropology reveals, however, in the absence of state-based or large-scale institutions, local leaders are often tasked with both managing various economic systems while also providing social services and interacting with followers on an interpersonal level [@cohen_comparative_1967; @fried_evolution_1967]. Within the domain of leader management, there are likely to be two primary features of human leader functions: strategizing and organizing. Our results correspond very closely to what @van_vugt_cognitive_2007 define as *strategic leadership*, where leaders increase the pay-offs of another individuals behavior thereby incentivizing followership versus *coordinating leadership*, where individuals bound by group challenges must effectively organize collective effort.

@@

The feature of leader organizational functions also aligns with the many evolutionary theories of leadership that emphasize the role of leaders in organizing collective actions [e.g., @hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; for review, see @garfield_evolutionary_2019]. In the ethnographic record such functions includes controlling and facilitating economic productivity and ensuring collective actions do not overly conflict with individual or household-level goals, such as among the Tukano of the Brazilian Amazon, where @sq19-018 [p. 127] notes:

@@

<!-- PROSCIAL FUNCTIONS FEATURE -->
**Public goods**. Beyond strategically shaping inter-individual behaviors or organizing collective behavior, we found much of leadership involves beneficial investments in group welfare. The prosocial feature of leader functions included distributing resources, providing protection, and various prosocial investments, which are functions commonly highlighted in the anthropological literature on leadership in nonindustrial societies [@sahlins_poor_1963; @johnson_organizational_1982; @service_origins_1975; @boehm_egalitarian_1993; @boehm_hierarchy_1999; @johnson_evolution_1987]. In the absence of institutionalized leadership roles, which characterize many of the relatively small-scale populations in this sample, leaders often maintain their social influence via continued community investment and the demands of followers can shape the functions of leaders to meet local needs.

@@

This study is based on the ethnographic record. Our methods are designed to capture a broad range of ethnographic descriptions from a culturally representative, systematic sample. Nonetheless, these results are conditional upon the content ethnographers chose to discuss and the way they chose to discuss it. The ethnographic record is certainly an imperfect source of information on human cultural diversity and hence a limitation of the current results. We nevertheless view the existence of the ethnographic record and the availability and functionality of the eHRAF as an enormous benefit for comparative research, given many of the populations represented in these data no longer exist or live as they did during the time of much ethnographic description. Within these ethnographic descriptions there is a general male bias, which could reflect a male bias in community leadership or the more systemic male bias in the ethnographic record [see @rosaldo_woman_1974; @mukhopadhyay_anthropological_1988]. These data are also biased towards cultural models (rather than descriptions of observed behavior) and community leadership; descriptions of kin group and residential group leadership are less frequent. The high proportion of text records discussing political group leaders could be attributable to our broad search strategy or the content of the ethnographic record of "leadership" more generally. Our sample of cultures and text records by continental region and subsistence strategy is also unbalanced.

Another limitation lies in our ability to interpret the absence of evidence. We are confident that the cross-culturally frequent measures discussed here and strongly represented in the ethnographic record and hence very likely to represent broad cross-cultural patterns in human behavior and culture. We cannot strongly conclude, however, that the relatively infrequent measures identified here are truly infrequent dimensions of human behavior and culture, only that they are relatively less frequent in the ethnographic record. Nonetheless, results reported here provide a broad, comprehensive, and cross-cultural framework for future empirical and theoretical work to draw on.

STATS limitations.

@@

# Concluding remarks: Beyond the non-WEIRD 
<!-- NOT SURE OF TITLE -- WANT TO EMPHASIZE NON WEIRD DIVERSITY AND ALSO MENTION BACK TO WERID FINDINGS BRIEFLY -->
In recognizing the limited generalizability of most extant psychological, sociological, and managerial data [@henrich_weirdest_2010], many social scientists are now diversifying their samples to better understand *human* psychology, behavior, organizations, and culture. We caution that the WEIRD people problem in behavioral science cannot be remedied simply by a "non-WEIRD" solution. Our results, and much of anthropology and ethnography, clearly demonstrate enormous diversity within populations that are now often categorized as "non-WEIRD", "traditional", or "small-scale" [e.g., @reyes-garcia_small-scale_2017]. Furthermore, higher-order explanatory variables such as "subsistence strategy" are probably less relevant for understanding the complexity of behavioral diversity than more salient group contexts. Despite limitations, the ethnographic record presents an invaluable opportunity to explore patterns of human cultural and behavioral diversity. Evolutionary theorists and researchers looking beyond WEIRD samples should take special interest in comparative ethnographic methods.

We found leaders in the ethnographic record are portrayed as generally prosocial, are equipped with special capacities, serve to resolve conflicts and organize cooperation, and receive material, social, and reproductive benefits. These results generally provide broad cross-cultural support for several influential theories, such as the dominance-prestige model, the service-for-prestige model, the computational services model, and numerous collective action models [for review, see @garfield_evolutionary_2019].

Cross-cultural evidence supports a dominance style of leadership involving coercive authority, confidence and aggression, as well as a prestige style involving special knowledge, experience, and innovation. We suggest, however, moving beyond dichotomous models of leadership strategies and focusing on inter-individual variation in broader typologies of leadership competencies. There is likely to be substantial overlap in, and interaction between physical, supernatural, cognitive, and other prosocial competencies. The degree to which a leader relies on such competencies will be shaped by their individual levels of them, the competencies of potential rivals, as well as their functions and group context, follower preferences, and cultural norms. For example, widespread preferences for egalitarianism and social equality are likely strong drivers of followers preferences for leaders across diverse societies [e.g., @cheng_dominance_2019; @price_evolution_2014], but shamanism and ritual based contexts represent a social domain in which ideologies support a style of leadership fusing both dominance and prestige-based competencies. Cultural forces and ideologies play important roles in shaping leadership styles and follower preferences including among WEIRD populations and in national political movements around the world [e.g., @mcdermott_leadership_2018; @inglehart_trump_2016] and the ecology of leadership dimensions can be better captured than dichotomous frameworks offer.

One interesting pattern that has not been widely discussed in the evolutionary literature on leadership is the role of shamans and their intersection, in our results, with the dominance and prestige styles of leadership. Our competencies cluster (Figure \@ref(fig:cluster)A) includes the *Supernatural* qualities dimension. @winkelman1987cross found that in societies sampled from the HRAF ($n = 43$) shamans provided decision-making services within judiciary, economic, military, or political domains in 44% or more of societies, and held charismatic leadership roles in 19% of societies; @winkelman1992shamans emphasizes the influential positions of leadership shamans attain through their charisma, social unification, healing abilities, and use of supernatural powers to cause harm. Providing ritual functions and military leadership are common (documented in over 50% of societies here) functions leaders serve (see Figure \@ref(fig:varsupport)).

>>>>

@@

```{r shamanism, fig.width=10, fig.cap="Coefficients of logistic elasticnet regression models of evidence for *Shamanism*. A: Coefficients indicate the words whose frequencies in each text record best predicted evidence for *Shamanism* in each text record. B: Non-zero coefficients of all quality and function variables that best predicted evidence for *Shamanism*. Both models used the lasso penalty ($\\alpha=1$) and chose $\\lambda_{1se}$ by cross-validation (1 SE from $\\lambda_{min}$). Red values are positive predictors; purple values are negative predictors. Variables with coefficents = 1 not displayed. X-axes are odds ratios on a log 10 scale."}
plot_shamanism_text <- plot_shamanism_text + scale_x_log10()

@@

We assessed the benefits and costs of leadership in two ways. First, we explicitly coded each text record for evidence of benefits and costs. Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported, suggesting leadership to be a high-risk, high-return strategy (see Figure \@ref(fig:varsupportcb)). Nevertheless, for most dimensions of benefits and costs, there was more evidence of benefits than costs (Figure \@ref(fig:benefitcostratio)), and the same was true averaging across all benefit/cost dimensions ($OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$). Second, several positive leader qualities indicating social, reproductive, and material success clustered together (bottom cluster Figure \@ref(fig:cluster)A). Both results, while not providing evidence of net fitness benefits, are nevertheless encouraging, especially since leaders had more evidence of mating and territory benefits than costs, which are key contributors to fitness, and also had more evidence for inclusive fitness, material and social benefits, relative to costs. This is consistent with much of the theoretical literature that links leader benefits to their capacities to provide prosocial benefits (e.g., resolving conflicts, organizing collective actions, and being generous) and/or impose social costs (e.g., punishing and aggressiveness) [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013; @hooper_theory_2010; @gavrilets_solution_2014; @price_evolution_2014; @smith_leadership_2016].

Among the Trobriand horticulturalists in Papua New Guinea, @ol06-037 [p. 47] describes relationships between polygynous marriages, increased wealth, and widespread social contacts: 
>Chiefs managed unusually large quantities of resources and it is agreed that the basis of such wealth was polygamy. Rank was quite directly expressed in the number of wives. The chief of Omarakana had 16 wives in Malinowski's time, while his predecessor Enamakala had 19 in 1893 (Annual Report 1893-4:19). In 1950 despite long-established missionary activity, there were still 13 (Powell 1960: 136). In traditional Trobriand society it seems that all leaders of village clusters of guyau status became polygamists. An emerging leader might also expect to be given wives by traditional allies outside the cluster while he might demand them from traditional rivals of subordinate subclans (Powell 1960:136)....

The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Followers were not reported to receive mating benefits, relative to leaders or other benefit types. For both leaders and followers, individuals stood to lose what they were able to gain. See Figure \@ref(fig:varsupportcb). @nm09-056 [p. 183] describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous female leader among the Iroquois, whose life exemplified the risks to both leaders and followers:

@@

@@

Shamanism appears to be a distinct form of leadership that combines a strategy of inducing fear, similar to the dominance strategy, but is based on knowledge and expertise, similar to the prestige strategy. The *Competencies* feature included supernatural and feared dimensions (Figure \@ref(fig:cluster)). Our *Shamanism* variable (which was not one of our original leadership dimensions) was predicted by words that evoke dominance and prestige, such as *supernatural*, *power*, *fear*, and *knowledge* (Figure \@ref(fig:shamanism)A), and by four other leadership dimensions that also overlap with the dominance and prestige constructs: *Feared*, *Medicinal functions*, *Ritual functions*, and *Experience/accomplished*. Shamans often provide valued public goods, have special expertise and abilities, are feared, and achieve or maintain influence via threats or use of direct and/or indirect aggression. @winkelman1987cross found that in societies sampled from the HRAF ($n = 43$) shamans provided decision-making services within judiciary, economic, military, or political domains in 44% or more of societies, and held charismatic leadership roles in 19% of societies; @winkelman1992shamans emphasizes the influential positions of leadership shamans attain through their charisma, social unification, healing abilities, and use of supernatural powers to cause harm. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate superhuman abilities [@singh_cultural_2017].

@ng06-054 describes shaman leaders among the Ojibwa:

@@

<!-- DISCUSSES BOTH MANAGEMENT FEATURES -->
**Leaders are managers**. There is a history within managerial studies among WEIRD populations of contrasting *leaders* from *managers*. @zaleznik_managers_1977, for example, suggested "managers" seek systematicity, organization, and stability, whereas "leaders" are willing to accept a lack of control and creatively navigate uncertainty to produce novel insight and appeal to the emotions of followers. Political anthropology reveals, however, in the absence of state-based or large-scale institutions, local leaders are often tasked with both managing various economic systems while also providing social services and interacting with followers on an interpersonal level [@cohen_comparative_1967; @fried_evolution_1967]. Within the domain of leader management, there are likely to be two primary features of human leader functions: strategizing and organizing. Our results (Figure \@ref(fig:cluster)B) correspond very closely to what @van_vugt_cognitive_2007 define as *strategic leadership*, where leaders increase the pay-offs of another individual's behavior thereby incentivizing followership, versus *coordinating leadership*, where individuals bound by group challenges must effectively organize collective effort.

@@

The feature of leader organizational functions aligns with the many evolutionary theories of leadership that emphasize the role of leaders in organizing collective actions [e.g., @hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; @pietraszewski_evolution_2019; for review, see @garfield_evolutionary_2019]. In the ethnographic record such functions includes controlling and facilitating economic productivity and ensuring collective actions do not overly conflict with individual or household-level goals, such as among the Tukano of the Brazilian Amazon, where @sq19-018 [p. 127] notes:

@@

<!-- PROSCIAL FUNCTIONS FEATURE -->
<!-- DISCUSSES PROSOCIALITY FEATURE -->
**Prosociality**. The prosocial cluster (Figure \@ref(fig:cluster)A), comprising prosocial competencies such as generosity, interpersonal and oratory skills, and charisma, and cultural conformity, such as proper behavior and having a favorable personality, generally corresponds to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @grabo_charisma_2017; @howell_role_2005]. The importance of leader prosociality also supports the @henrich_big_2015 Big-man mechanism by which the positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

We also found that leader functions often involve prosocial investments in group welfare (Figure \@ref(fig:cluster)B). The prosociality feature of leader functions included distributing resources, providing protection, and various prosocial investments, which are functions commonly highlighted in the anthropological literature on leadership in nonindustrial societies [@sahlins_poor_1963; @johnson_organizational_1982; @service_origins_1975; @boehm_egalitarian_1993; @boehm_hierarchy_1999; @johnson_evolution_1987]. In the absence of institutionalized leadership roles, which characterize many of the relatively small-scale populations in this sample, leaders often maintain their social influence via continued community investment and the demands of followers can shape the functions of leaders to meet local needs.

## The six-fold way

In our view, the dominance-prestige and "dual-process" models of leadership, while advancing the discipline, cannot easily account for shaman leaders nor "managers". In short, these models have conflated "benefits" with knowledge and expertise, and "costs" with physical formidability. Instead, as shaman leaders demonstrate, knowledge and expertise can both provide benefits and impose costs, and the same goes for physical formidability. The "dual" in dual process comes from the fact that one individual can influence others either by providing benefits or by imposing costs. Moreover, not all leaders have status (e.g., managers), and not all high status individuals are leaders. We therefore tenatively advance an alternative scheme whereby leaders possess some combination of knowledge/expertise and physical formidability which they use to provide benefits or impose costs. If their abilities to provide benefits or impose costs are exceptional relative to other group members, they gain status. Hence, we should be able to find examples of leaders with various combinations of these five elements.

In our data, we combinded a subset of our leader dimensions into these five elements. We then calculated the fraction of text records with evidence for one or more of the five elements. See Figure \@ref(fig:fivefold).

```{r fivefold, fig.width=10, fig.cap=""}
upset(df_fivefold, order.by='freq', nintersects = 19, text.scale = 2)
```

@@

The ethnographic record is an imperfect source of information on human cultural diversity. Our results are conditional upon the content ethnographers chose to discuss, the way they chose to discuss it, and the way we interpreted and coded their texts. There was a general male bias, which could reflect a male bias in community leadership or a more systemic male bias in the ethnographic record [see @rosaldo_woman_1974; @mukhopadhyay_anthropological_1988]. These data are also biased towards cultural models of leadership (rather than particular cases of leadership) and community leadership; descriptions of kin group and residential group leadership were less frequent. The high proportion of text records discussing political group leaders could be attributable to our broad search strategy or the content of the ethnographic record of "leadership" more generally. Our sample of cultures and text records by continental region and subsistence strategy is also unbalanced. We also detected bias in evidence by date of publication for nine of our 109 leader dimensions (see Figure S\@ref(fig:pubdate)).

Another limitation lies in our ability to interpret the absence of evidence. We are confident that the cross-culturally frequent measures discussed here are strongly represented in the ethnographic record and hence very likely to represent broad cross-cultural patterns. We cannot strongly conclude, however, that the relatively infrequent measures identified here are truly infrequent dimensions of human behavior and culture, only that they are relatively less frequent in the ethnographic record.

This is an exploratory study, which means that our analyses were driven by the data rather than by *a priori* hypotheses. This has the advantage that unexpected patterns can be discovered, but the disadvantage that some random variation in the data can, and almost certainly will, be misinterpreted as meaningful patterns. In particular, our cluster analysis of leader qualities (Figure \@ref(fig:cluster)A) was sensitive to the choice of different clustering algorithms and inclusion or exclusion of variables with little evidence.

@@

# Concluding remarks: Beyond the non-WEIRD 
<!-- NOT SURE OF TITLE -- WANT TO EMPHASIZE NON WEIRD DIVERSITY AND ALSO MENTION BACK TO WERID FINDINGS BRIEFLY -->
In recognizing the limited generalizability of most extant psychological, sociological, and managerial data [@henrich_weirdest_2010], many social scientists are now diversifying their samples to better understand *human* psychology, behavior, organizations, and culture. We caution that the WEIRD people problem in behavioral science cannot be remedied simply by a "non-WEIRD" solution. Our results, and much of anthropology and ethnography, clearly demonstrate enormous diversity among populations that are now often categorized as "non-WEIRD", "traditional", or "small-scale". At the same time, anthropologists have made the opposite mistake in often failing to recognize deep similarities among peoples of all cultures [@brown_human_1991; @said1979; @reyes-garcia_small-scale_2017].

Our results support both important dimensions of leadership that could be human universals, as well as systematic variation in other dimensions. We found leaders in the ethnographic record are portrayed as generally prosocial, are equipped with special competencies, serve to resolve conflicts and organize cooperation, and receive material, social, and reproductive benefits. Cross-cultural evidence also supports a dominance style of leadership involving coercive authority, punishment, aggression, and control of resources. These results generally provide broad cross-cultural support for several influential theories, such as the dominance-prestige model, the service-for-prestige model, the computational services model, and numerous collective action models [for review, see @garfield_evolutionary_2019].

Shamans emerged as an important category of leaders that are rarely discussed in the evolutionary literature [but see @singh_cultural_2017]. Supernatural qualities were among the top ten leader qualities across cultures (Figure \@ref(fig:varsupport)). Shamans share qualities with prestige-style leaders, such as having specialized knowledge and abilities, yet appear, at least at times, to use their knowledge to instill fear, similar to dominance-style leaders.

[@hawley_ontogenesis_1999; @hawley_myth_2008]

Leaders vs. Status vs. Provide benefits vs. Impose costs vs. Information vs. Physical

We therefore suggest moving beyond dichotomous, or "dual-process", models of leadership.

and focusing on inter-individual variation in broader typologies of leadership competencies. There is likely to be substantial overlap in, and interaction between physical, supernatural, cognitive, and other prosocial competencies. The degree to which a leader relies on such competencies will be shaped by their individual levels of them, the competencies of potential rivals, as well as their functions and group context, follower preferences, and cultural norms. For example, widespread preferences for egalitarianism and social equality are likely strong drivers of followers preferences for leaders across diverse societies [e.g., @cheng_dominance_2019; @price_evolution_2014], but shamanism and ritual based contexts represent a social domain in which ideologies support a style of leadership fusing both dominance and prestige-based competencies. Cultural forces and ideologies play important roles in shaping leadership styles and follower preferences including among WEIRD populations and in national political movements around the world [e.g., @mcdermott_leadership_2018; @inglehart_trump_2016] and the ecology of leadership dimensions can be better captured than dichotomous frameworks offer.

Furthermore, higher-order explanatory variables such as "subsistence strategy" are probably less relevant for understanding the complexity of behavioral diversity than more salient group contexts.
